Omics-based Single Step Trait Prediction

### 2017-10-08 paper/manuscript.tex
paper, add AEM review from 2017-09-20

\begin{abstract}
Predicting genotypic values with great precision is pivotal for effective
candidate selections in animal and plant breeding.
Novel `omics' predictors have shown to improve upon established genome-based
predictions of some complex traits but are yet expensive to sample.
In animal breeding, single step prediction was introduced to efficiently use
pedigree information, collected on a large number of animals, and genomic
information, collected on a smaller subset of animals, simultaneously without
bias.
Single step prediction has not been attempted with a predictor whose features
display heritabilities smaller than one, yet.
Using two maize datasets of inbred lines and hybrids we show that transcriptomic
data can be imputed successfully, boosting forecasts when their predictive
ability exceeds that of pedigree or genomic data.
Covering only a subset of genotypes with `omics' predictors and imputing all
others using pedigree or genomic data could enable breeders to improve trait
predictions while keeping costs manageable.
Employing `omics' predictors could particularly benefit candidate selection in
hybrid breeding where the success of forecasts is a strongly convex function of
predictive ability.
\end{abstract}

@@

Recent studies in maize inbred lines~\cite{Guo2016}, maize hybrids
\cite{Zenke-Philippi2017,Westhues2017}, rice hybrids~\cite{Dan2016,Xu2016} and
humans~\cite{Vazquez2016} have suggested that `omics' predictors, such as the
metabolome and the transcriptome, can improve the prediction of traits with
complex genetic architectures over that achieved with pedigree or genomic data.
Particularly hybrid breeding, where the success of forecasts increases
exponentionally with predictive abilities, could benefit from the use of
predictors that yield more precise estimates of candidate performance.
Currently, the adoption of these novel predictors by breeders might be impeded
by their higher sampling costs compared to pedigree and genomic data.
Here, we present single step prediction of complex traits as a framework for
combining pedigree and/or genomic data with `omics' predictors collected on a
subset of all available genotypes to boost predictive abilities while managing
costs.

@@

\begin{enumerate}
\item In both data sets, single step prediction improved predictive abilities
	over those of an individual predictor when the discrepancy between the
	two predictors in terms of predictive ability were moderate to large in
	favor of the incomplete predictor.
\item When 20\%-30\% ($n \approx 25$) of the genotypes were covered by both
	predictors, single step-based predictive abilities converged to those
	obtained with the incomplete predictor.
\end{enumerate}

In conclusion our results demonstrate that single step prediction, using
an established predictor such as pedigree information, can benefit from sampling
a highly performant predictor, such as transcriptomic data, on a subset of all
genotypes.
The framework presented herein should help promote the use of attractive
`omics' predictors by balancing their advantages with their currently higher
sampling costs.

% Introduction
\section{Introduction}
% Rationale for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
animal and plant breeding by a more efficient selection of promising candidates 
without phenotypes~\cite{DeLosCampos2013,Garcia-Ruiz2016}.
Its use is largely motivated by the high costs and oftentimes long duration of 
phenotyping, which can be reduced when all genotypes are equipped with a predictor 
while limiting phenotyping to a few genotypes with high precision
\cite{Kadam2016}.
The performance of all other genotypes is then forecast using model parameters
estimated in the subset of individuals with full information.
In animal breeding, genomic selection has increased the annual selection gain 
by $\approx$ 50--100\% for yield traits and by 300--400\% for traits with low 
heritability~\cite{Garcia-Ruiz2016}.

% Incomplete predictors.
Breeding programs --- particularly in animal breeding --- are characterized by
the availability of one predictor type, which is available for all recorded
genotypes while other predictors are incomplete in that they cover only a 
subset of all genotypes~\cite{Fragomeni2015}.
Usually, the complete predictor is pedigree information while genomic data are
the most prevalent incomplete predictor.
To utilize all available phenotypic information for model training, methods were 
developed that allow the combined use of genomic and pedigree information 
\cite{Hayes2009a,VanRaden2009}.

% Two-step prediction.
Initially, a two-step procedure was the most intuitive choice for accomplishing
this goal~\cite{VanRaden2009}.
The first step consists of conventionally estimating breeding values based on
pedigree relationships and the phenotypes of relatives.
Individuals with accurate breeding values out of the first step, which have a
high number of offspring, enter the second step in which conventional breeding
values are regressed onto SNP genotypes.
The estimated model coefficients can then be used to predict breeding values
for solely genotyped and potentially very young individuals.
However, as the breeding values from step one enter the genomic prediction
model as phenotypes, one has to be careful about the residual error
distribution of such a model.
For instance, the range of the prediction errors will no longer be identically
distributed and thus, they should be weighted differently or estimated 
individually~\cite{Aguilar2010}.
One popular approach to account for this problem is deregressing the breeding
values and weighting the residuals according to the prediction error variances
\cite{Garrick2009}.

% Workings and advantages of the single step procedure.
Independently from each other,~\citeNP{Legarra2009} and
\citeNP{Christensen2010} developed a single step BLUP method that blends the 
numerator relationship matrix $\mathbf{A}$ with the genomic relationship matrix
$\mathbf{G}$ in a mutual matrix $\mathbf{H}$ to use all predictor information
simultaneously.
Issues with this H-BLUP approach were the compatability of $\mathbf{A}$ with
$\mathbf{G}$~\cite{Christensen2012} and their weighting, which is not
trait-independent~\cite{Vitezica2011,Ashraf2016}.
\citeN{Fernando2014} derived the equivalent single step marker-effect-model to 
the single step breeding value model of \citeNP{Legarra2009} and 
\citeNP{Christensen2010}, which does not require weights for $\mathbf{A}$ and 
$\mathbf{G}$ and allows implicit modeling of the imputation error structure.

% -   Hybrids
In animal breeding, single step prediction, particularly H-BLUP, has already 
become a routine procedure~\cite{Legarra2014} but, to the best of our knowledge, 
has only been considered in plants in a study on wheat~\cite{Ashraf2016} and
white spruce lines, respectively~\cite{Ratcliffe2017}.
Moreover, while an application of the single step framework to animals exists
for pigs~\cite{Xiang2016}, it is yet lacking for plant hybrids.
Technological advances in plant hybrid breeding such as the rapid generation
of completely homozygous lines~\cite{Wedzony2009} now enable breeders to
produce about $n = 1,000$ individuals per generation and sex.
Assuming that about 90\% of these lines were never before tested in any hybrid
combination about $n^{2} * 0.9^{2} = 810,000$ putative hybrids per generation
would not have a single parent with phenotypically tested progeny
\cite{Westhues2017}.
Predicting these hybrid candidates with low relatedness to tested hybrids using
only information from their parents is currently a hot topic in plant breeding
\cite{Technow2014,Kadam2016,Westhues2017}.

% -   quantitative predictores
Given that the availability of pedigree and genomic data is the most prevalent
situation in animal breeding, previous studies have focused exclusively on the
application of single step prediction to these predictors.
Recent investigations into the utility of `omics' data for boosting predictive
abilities in plants have suggested that metabolites and transcriptomic data can
improve upon predictive abilities based on pedigree or genomic data
\cite{Guo2016,Xu2016,Dan2016,Zenke-Philippi2017,Westhues2017}.
Yet, these technologies are not fully mature and therefore incur higher costs
to breeders than pedigree and genomic data, putting their utility into question
despite their proven usefulness.

% - Objectives
Hence, our objectives were to i) examine whether the single step prediction 
framework can be transfered to a situation where a very large number of hybrids
($\approx$ 1 million) needs to be predicted and ii) evaluate the impact of
covering only a subset of all genotypes with both, the complete and the
incomplete predictor on predictive ability, where for the remaining individuals
only genomic data are available.

\section{Materials and Methods}
\Genetics2level{Exp1}
To address the second objective a maize inbred line data set, subsequently
referred to as Exp1, was used~\cite{Yang2014}.
Exp1 originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four subgroups classified by
\citeN{Guo2016}.
All inbred lines were evaluated in five Chinese environments described in detail
by~\citeNP{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated by~\citeN{Yang2014} for
17 traits of which six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that two traits were predicted with
greater precision by transcriptomic data and four further traits were predicted
with greater precision by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of 513 lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploiting identity by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
for 37,760 SNPs using the Beagle software~\cite{Browning2009}.
Gene expression data were normalized using a normal quantile transformation to
satisfy modeling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes~\cite{Fu2013} and provided the basis for further analyses.

\Genetics2level{Exp2}
To address objectives 1 and 2, a maize hybrid data set, subsequently
denoted as Exp2, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines described in detail by
\citeN{Westhues2017}.
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse environments across Germany for seven
agronomically important traits in silage maize by accounting for year, location,
field replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
The same SNP quality checks as for the material in Exp1 were applied here,
yielding 7,013 polymorphic marker loci for the Dent and 6,212 for the Flint
lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 Dent and 43 Flint lines, that are parents of 685 hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts
\cite{Westhues2017} using the R-package \emph{limma}~\cite{Ritchie2015a}.

@@

\Genetics2level{Population structure}
Population structure in Exp1 was examined through a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and carried out a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the \emph{R}-package
\emph{LEA}~\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \emph{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To mitigate the possibility of inflated predictive abilities, which can arise
from large differences in the average performance of the different
subpopulations~\cite{Windhausen2012}, we reduced the panel of inbred lines to
those assigned to the largely overlapping clusters 1 and 4 in
Fig.~\ref{fig:inbred-pca}.
This reduction yielded 164 genotypes, 110 of which were covered by both, genomic
and transcriptomic data.

@@

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, and $\mathbf{e}$ is the 
residual error.
For the inbred lines we fit random genotype effects $\mathbf{g}$.
For both experiments, genetic effects $\mathbf{g}$ are associated with the
vector of phenotypic performance $\mathbf{y}$ through a corresponding design
matrix $\mathbf{Z}$.
The random effects ($\mathbf{g}$) have expectation zero and variance equal to
$\mathbf{K} \sigma^{2}_{g}$ and $\mathbf{I} \sigma^2_{e}$ for the residual
error.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
``omics'' predictors in the same material concluded no considerable gains 
\cite{Westhues2017}.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

\subsection{Single step prediction}
If the breeding values are fully captured by the features in $\mathbf{M}$,
Eq.~\ref{eq:KBLUPModel} can be expressed as

@@

\cite{Fernando2014} because the ridge regression model is equivalent to the
BLUP model~\cite{Ruppert2003}.
In the previous equation, $\mathbf{M}$ is a matrix of feature covariates and
$\boldsymbol{\alpha}$ is the vector of random, partial-regression coefficients
of the feature covariates.

Hence, the breeding values in a prediction model are given by:
\begin{equation} \label{eq:mrnaebv}
	\mathbf{\hat{g}} = \mathbf{M}\boldsymbol{\hat{\alpha}},
\end{equation}

@@

Consider now the situation in which one predictor is complete in the sense that
the whole population is equipped with features for that predictor whereas only
a subset of that population has information on a second predictor.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor by using covariates of the complete predictor.
Let the subscript $1$ denote individuals covered only by the complete
predictor whereas individuals covered by both, the complete and the
incomplete predictor, are indicated by the subscript $2$.
Let the index $C$ denote matrices and vectors pertaining to the complete
predictor and let the index $I$ denote matrices and vectors pertaining to the
incomplete predictor, respectively.

Let us further denote the matrix of centered covariates of the incomplete
predictor as
\begin{equation} \label{eq:incomplete-covariates}
\tensor[_{I}]{\mathbf{W}}{} =
\begin{bmatrix}
  \mathbf{Z}_{1} \; \tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} \\
  \tensor[_I]{\mathbf{W}}{_{2}}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\tensor[_I]{\mathbf{W}}{_{2}}$
while the feature covariates for genotypes not covered by the incomplete
predictor can be estimated/imputed with the aid of the covariance matrix of the
complete predictor $\tensor[_{C}]{\mathbf{K}}{}$ and the feature matrix of the
incomplete predictor $\tensor[_{I}]{\mathbf{W}}{_{2}}$ by calculating
($\tensor[_{I}]{\mathbf{\hat{W}}}{_{1}} = \tensor[_C]{\mathbf{K}}{_{12}} \;
\tensor[_C]{\mathbf{K}}{_{22}^{-1}} \; \tensor[_I]{\mathbf{W}}{_{2}}$).
In this case we need to partition the vector of genetic values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} genetic values of genotypes
covered only by the complete predictor) and a component
$\mathbf{g}_{2} = \tensor[_{I}]{\mathbf{W}}{_{2}} \; 
\tensor[_{I}]{\boldsymbol{\alpha}}{}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors where $\tensor[_{I}]{\boldsymbol{\alpha}}{}$
pertains to the random feature effects of the incomplete predictor).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) +
	\boldsymbol{\epsilon} \\
	&= \tensor[_{C}]{\mathbf{K}}{_{12}} \;
       	\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[_{I}]{\mathbf{W}}{_{2}} \;
	\tensor[_{I}]{\boldsymbol{\hat{\alpha}}}{} +
       	(\mathbf{g_1} - \tensor[_{C}]{\mathbf{K}}{_{12}} \;
	\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[_{I}]{\mathbf{W}}{_{2}} \;
	\tensor[_{I}]{\boldsymbol{\hat{\alpha}}}{}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

where $\tensor[_{C}]{\mathbf{K}}{_{ij}}$ are partitions of the matrix
$\tensor[_{C}]{\mathbf{K}}{}$ that correspond to $\mathbf{g}_{1}$ and
$\mathbf{g}_{2}$, respectively~\cite{Fernando2014} and where
$\mathbf{\hat{g}}_{1}$ pertains to the imputed genetic values of genotypes
not equipped with the incomplete predictor.

The covariance matrix of $\boldsymbol{\epsilon}$ is
$\tensor[_{C}]{\mathbf{K}}{_{11}} -
\tensor[_{C}]{\mathbf{K}}{_{12}} \;
\tensor[_{C}]{\mathbf{K}}{_{22}^{-1}} \;
\tensor[_{C}]{\mathbf{K}}{_{21}} =
(\tensor[_{C}]{\mathbf{K}}{^{11}})^{-1}$
\cite{Legarra2009}

>>>>

\begin{abstract}
Predicting genotypic values with great precision is pivotal for effective
candidate selections in animal and plant breeding.
Novel `omics' predictors have shown to improve upon established genome-based
predictions of some complex traits but yet require laborious and expensive
assays.
In animal breeding, single-step prediction was introduced to efficiently use
pedigree information, collected on a large number of animals, and genomic
information, collected on a smaller subset of animals, simultaneously for breeding
value estimation without bias.
Single-step prediction has not yet been attempted with predictors calculated from
quantitative features displaying heritabilities smaller than one.
Using two maize datasets of inbred lines and hybrids we show that transcriptomic
data can be imputed successfully, boosting forecasts when their predictive
ability exceeds that of pedigree or genomic data.
Covering only a subset of genotypes with `omics' predictors and imputing all
others using pedigree or genomic data could enable breeders to improve trait
predictions while keeping costs under control.
Employing `omics' predictors could particularly improve candidate selection in
hybrid breeding where the success of forecasts is a strongly convex function of
predictive ability.
\end{abstract}

@@

Recent studies in maize inbred lines~\cite{Guo2016}, maize hybrids
\cite{Zenke-Philippi2017,Westhues2017}, rice hybrids~\cite{Dan2016,Xu2016} and
humans~\cite{Vazquez2016} have suggested that `omics' predictors, such as the
metabolome and the transcriptome, can improve the prediction of traits with
complex genetic architectures over that achieved with pedigree or genomic data.
Particularly hybrid breeding, where the success of forecasts increases
exponentionally with predictive abilities, could benefit from the use of
predictors that yield more precise estimates of candidate performance.
Currently, the adoption of these novel predictors by breeders is impeded
by their higher sampling costs compared to pedigree and genomic data.
Here, we present single-step prediction of complex traits as a framework for
combining pedigree and/or genomic data with other `omics' predictors collected
on a subset of all available genotypes to boost predictive abilities while
managing costs.

@@

\begin{enumerate}
\item In both data sets, single-step prediction improved predictive abilities
	over those of the predictor covering all genotypes when the discrepancy 
	between the	two predictors in terms of predictive ability were moderate to
	large in favor of the incomplete predictor.
\item When 20\%-30\% ($n \approx 25$) of the genotypes were covered by both
	predictors, single step-based predictive abilities converged to those
	obtained with the incomplete predictor.
\end{enumerate}

In conclusion our results demonstrate that single-step prediction, using
an established predictor such as pedigree information, can benefit from sampling
a highly performant predictor, such as transcriptomic data, on a subset of all
genotypes.
The framework presented herein should help promote the use of attractive
`omics' predictors by balancing their advantages with their currently higher
costs.

% Introduction
\section{Introduction}
% Rationale for predictions.
Genomic prediction, pioneered by \citeNP{Meuwissen2001}, has revolutionized
animal and plant breeding by a more efficient selection of promising candidates 
without phenotypic records~\cite{DeLosCampos2013,Garcia-Ruiz2016}.
Its use is mainly motivated by the high costs and oftentimes time-consuming
phenotyping, which can be reduced when all genotypes are covered by a predictor 
while limiting phenotyping to a few genotypes with high precision
\cite{Kadam2016}.
The performance of all other genotypes is then forecast using model parameters
estimated in the subset of individuals with full information.
In animal breeding, genomic selection has increased the annual selection gain 
by $\approx$ 50--100\% for yield traits and by 300--400\% for traits with low 
heritability~\cite{Garcia-Ruiz2016}.

% Incomplete predictors.
Breeding programs --- particularly in animal breeding --- are characterized by
the availability of one predictor type, which is available for all recorded
genotypes with phenotypic information while other predictors are incomplete in
that they cover only a subset of all genotypes~\cite{Fragomeni2015}.
Usually, the complete predictor is pedigree information while genomic data are
the most prevalent incomplete predictor.
To utilize all available phenotypic information for model training, methods were 
developed that allow the combined use of genomic and pedigree information 
\cite{Hayes2009a,VanRaden2009}.

% Two-step prediction.
Initially, a two-step procedure was the most intuitive choice for accomplishing
this goal~\cite{VanRaden2009}.
The first step consists of conventionally estimating breeding values based on
pedigree relationships among the candidates and phenotypes of their relatives.
Individuals with higly reliable breeding values out of the first step,
(\textit{i.e.} those having a high number of offspring) enter the second step in
which conventional breeding values are regressed onto SNP genotypes.
The estimated model coefficients can then be used to predict breeding values
for solely genotyped and potentially very young individuals.
However, as the breeding values from step one enter the genomic prediction
model as phenotypes, one has to be careful about the residual error
distribution of such a model.
For instance, the range of the prediction errors will no longer be identically
distributed and thus, they should be weighted differently or estimated 
individually~\cite{Aguilar2010}.
One popular approach to tackle this problem is deregressing the breeding
values and weighting the residuals according to the prediction error variances
\cite{Garrick2009}.

% Workings and advantages of the single step procedure.
\citeNP{Legarra2009} and \citeNP{Christensen2010} independently from each other
developed a single-step BLUP method that blends the numerator relationship matrix
$\mathbf{A}$ with the genomic relationship matrix $\mathbf{G}$ in a mutual matrix
$\mathbf{H}$ to use all predictor information simultaneously.
Issues with this H-BLUP approach were the compatability of $\mathbf{A}$ with
$\mathbf{G}$~\cite{Christensen2012} and their weighting, which is not
trait-independent~\cite{Vitezica2011,Ashraf2016}.
\citeN{Fernando2014} derived the equivalent single-step marker-effect-model to 
the single-step breeding value model of \citeNP{Legarra2009} and 
\citeNP{Christensen2010}, which does not require weights for $\mathbf{A}$ and 
$\mathbf{G}$ and allows implicit modeling of the imputation error structure.

% -   Hybrids
In animal breeding, single-step prediction, particularly H-BLUP, has already 
become a routine procedure~\cite{Legarra2014} for predicting breeding values
but, to the best of our knowledge, has only been considered in plants in a study
on wheat~\cite{Ashraf2016} and white spruce lines, respectively
\cite{Ratcliffe2017}.
Moreover, while an extension of the single-step framework to hybrid genotypes
was developed for pigs~\cite{Xiang2016}, it is yet lacking for the prediction
of single-cross hybrids between purely inbred lines.
Technological advances in hybrid breeding of several crop species, such as the
rapid generation of completely homozygous lines~\cite{Wedzony2009}, now enable
plant breeders to produce about $n = 1,000$ individuals per generation and sex.
Assuming that about 90\% of these lines were never before tested in any hybrid
combination about $n^{2} * 0.9^{2} = 810,000$ putative hybrids per generation
would not have a single parent with phenotypically tested progeny
\cite{Westhues2017}.
Predicting the performance of these hybrid candidates with low relatedness to 
tested hybrids using only information from their parents is currently a hot topic
in plant breeding~\cite{Technow2014,Kadam2016,Westhues2017}.

% -   quantitative predictores
Given that the availability of pedigree and genomic data is the most prevalent
situation in animal breeding, previous studies have focused exclusively on the
application of single-step prediction to these predictors.
Recent investigations into the utility of `omics' data for boosting predictive
abilities in plants have suggested that metabolic and transcriptomic data can
improve upon predictive abilities based on pedigree or genomic data
\cite{Guo2016,Xu2016,Dan2016,Zenke-Philippi2017,Westhues2017}.
Yet, these two `omics' technologies are not fully mature and therefore incur
higher costs to breeders than pedigree and genomic data, putting their utility
into question despite their proven usefulness.

% - Objectives
Hence, our objectives were to (i) extend the existing theoretical framework for
single-step prediction to the imputation of quantitative `omics' features, here
exemplified by transcriptomic data, (ii) evaluate the impact of the proportion of
imputed genotypes in the training set on the predictive ability and (iii) transfer
the single-step framework for breeding value prediction to the prediction of
hybrids.

\section{Materials and Methods}
\Genetics2level{Exp1}
To address objectives (i) and (ii) a published data set of maize inbred lines
\cite{Yang2014} was used.
It originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$), subsequently referred to as Exp1, which is the largest of the four
subgroups classified by~\citeN{Guo2016}.
All inbred lines were evaluated in five different environments in China
described in detail by~\citeNP{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated by~\citeN{Yang2014} for
all 211 inbred lines and for 17 traits of which six are presented in this study 
(Table~\ref{table:TraitKey}).
The selection of these traits was done based on our prediction results so that
two traits were predicted with greater precision by transcriptomic data and four
further traits were predicted with greater precision by genomic data.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed
using the Beagle software~\cite{Browning2009} resulting in 37,760 SNPs for these
211 genotypes.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes~\cite{Fu2013} and provided the basis for further analyses.

\subsection{Population structure}
Population structure in Exp1 was examined through a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated likelihood of the data for each $K$, we chose $K=4$
and carried out a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the \emph{R}-package
\emph{LEA}~\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \emph{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To mitigate the possibility of inflated predictive abilities, which can arise
from large differences in the average performance of the different
subpopulations~\cite{Windhausen2012}, we reduced the panel of inbred lines to
164 genotypes, which were assigned to the largely overlapping clusters 1 and 4 in
Fig.~\ref{fig:inbred-pca}.
In summary, genomic and phenotypic information were available for 164 inbred
lines, which we denote as the `Full' set whereas transcriptomic data were available
for a subset of 110 inbred lines, which we denote as `Reduced'.

\Genetics2level{Exp2}
To address objectives (i) and (iii), a maize hybrid data set, subsequently
denoted as Exp2, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines described in detail by
\citeN{Westhues2017}.
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse environments across Germany for seven
agronomically important traits in silage maize by accounting for year, location,
field replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
The same SNP quality checks as for the material in Exp1 were applied here,
yielding 7,013 polymorphic marker loci for the Dent and 6,212 for the Flint
lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 Dent and 43 Flint lines, that are parents of 685 hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts
\cite{Westhues2017} using the R-package \emph{limma}~\cite{Ritchie2015a}.

@@

@@

where $\mathbf{y}$ is the vector of observed inbred line performance $\mu$ is
the fixed model intercept, and $\mathbf{e}$ is the residual error.
For the inbred lines we fit random genotype effects $\mathbf{g}$, which are
associated with the vector of phenotypic performance $\mathbf{y}$ through a
corresponding design matrix $\mathbf{Z}$.
The random effects ($\mathbf{g}$) have expectation zero and variance equal to
$\mathbf{K} \sigma^{2}_{g}$ and $\mathbf{I} \sigma^2_{e}$ for the residual
error, respectively.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
``omics'' predictors in the same material concluded no considerable gains 
\cite{Westhues2017}.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

\subsection{Single-step prediction}
If the breeding values are fully captured by the features in $\mathbf{M}$,
Eq.~\ref{eq:KBLUPModel} can be expressed as~\cite{Fernando2014}

@@

with $\boldsymbol{\alpha}$ being the vector of random, partial-regression
coefficients of the feature covariates, because the ridge regression model is
equivalent to the BLUP model \cite{Ruppert2003}.

Hence, the breeding values from the prediction model are given by:
\begin{equation} \label{eq:mrnaebv}
	\mathbf{\hat{g}} = \mathbf{M}\boldsymbol{\hat{\alpha}},
\end{equation}

@@

Consider now the situation in which one predictor is complete in the sense that
the whole population is covered by features for that predictor but only a subset
of that population has information for a second predictor.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor, denoted as `predictor' and indicated through the
asterisk symbol `*', by using covariates of the complete predictor, denoted as
`imputor'.
Let the subscript $1$ denote individuals covered only by the predictor whereas
individuals covered by both, the imputor and the predictor, are indicated by the subscript $2$.

Let us further denote the matrix of centered covariates of the predictor as
\begin{equation} \label{eq:incomplete-covariates}
\tensor[]{\mathbf{W}}{^{*}} =
\begin{bmatrix}
  \mathbf{Z}_{1} \; \tensor[]{\mathbf{\hat{W}}}{^{*}_{1}} \\
  \tensor[]{\mathbf{W}}{^{*}_{2}}
 \end{bmatrix},
\end{equation}
where observed feature covariates are stored in $\tensor[]{\mathbf{W}}{^{*}_{2}}$
while the feature covariates for genotypes not covered by the predictor can be estimated/imputed with the aid of the covariance matrix of the
imputor $\mathbf{K}$ and the feature matrix of the predictor 
$\tensor[]{\mathbf{W}}{^{*}_{2}}$ by calculating
($\tensor[]{\mathbf{\hat{W}}}{^{*}_{1}} = \tensor[]{\mathbf{K}}{_{12}} \;
\tensor[]{\mathbf{K}}{_{22}^{-1}} \; \tensor[]{\mathbf{W}}{^{*}_{2}}$).
In this case we need to partition the vector of breeding values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} breeding values of genotypes
covered only by the imputor) and a component
$\mathbf{g}_{2} = \tensor[]{\mathbf{W}}{^{*}_{2}} \; 
\tensor[]{\boldsymbol{\alpha}}{^{*}}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors, where
$\tensor[]{\boldsymbol{\alpha}}{^{*}}$ pertains to the random feature effects of
the predictor).
Features are distributed according to a multivariate normal distribution:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) +
	\boldsymbol{\epsilon} \\
	&= \tensor[]{\mathbf{K}}{_{12}} \;
       	\tensor[]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[]{\mathbf{W}}{^{*}_{2}} \;
	\tensor[]{\boldsymbol{\hat{\alpha}}}{^{*}} +
       	(\mathbf{g_1} - \tensor[]{\mathbf{K}}{_{12}} \;
	\tensor[]{\mathbf{K}}{_{22}^{-1}} \;
       	\tensor[]{\mathbf{W}}{^{*}_{2}} \;
	\tensor[]{\boldsymbol{\hat{\alpha}}}{^{*}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

where the first summand pertains to $\mathbf{\hat{g}}_{1}$, the second summand
pertains to $\boldsymbol{\epsilon}$ and where the partitions of the matrix
\begin{equation} \label{eq:imputor}
\tensor[]{\mathbf{K}}{} =
\begin{bmatrix}
  \mathbf{K}_{11} & \mathbf{K}_{12} \\
  \mathbf{K}_{21} & \tensor[]{\mathbf{K}_{22}}{^{*}}
 \end{bmatrix},
\end{equation}

correspond to $\mathbf{g}_{1}$ and $\mathbf{g}_{2}$, respectively
\cite{Fernando2014} and $\mathbf{\hat{g}}_{1}$ pertains to breeding values
estimated by imputing the predictor.

The covariance matrix of $\boldsymbol{\epsilon}$ is
$\tensor[]{\mathbf{K}}{_{11}} -
\tensor[]{\mathbf{K}}{_{12}} \;
\tensor[]{\mathbf{K}}{_{22}^{-1}} \;
\tensor[]{\mathbf{K}}{_{21}} =
(\tensor[]{\mathbf{K}}{^{11}})^{-1}$
\cite{Legarra2009}

and the covariance of $\mathbf{\hat{g}}_{1}$ is
$\tensor[]{\mathbf{K}}{_{12}} \;
\tensor[]{\mathbf{K}}{_{22}^{-1}} \;
\tensor[]{\mathbf{W}}{^{*}_{2}} \;
\tensor[]{\mathbf{W}}{^{*}_{2}^{\top}} \;
\tensor[]{\mathbf{K}}{_{22}^{-1}} \;
\tensor[]{\mathbf{K}}{_{21}} \;
\tensor[]{\sigma^{2}_{\alpha}}{^{*}}$
\cite{Fernando2014}.

The structure of the residual imputation/prediction error is known, and
therefore, can be modeled.

The general model for the single-step procedure can then be written as

\begin{equation} \label{eq:single-step-model}
\mathbf{y} =
\mathbf{Xb} +
\tensor[]{\mathbf{W}}{^{*}} \;
\tensor[]{\boldsymbol{\alpha}}{^{*}} +
\mathbf{U} \boldsymbol{\epsilon} +
\mathbf{e},
\end{equation}
